---
title: "Checkpoint#04_Yihan Zhong"
author: "Yihan Zhong"
date: "2026-03-06"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(haven)
library(dplyr)
library(ggplot2)
library(knitr)
library(tidyr)
```

# Introduction

This report builds on previous checkpoints to conduct data simulation and power analysis in the context of examining the association between smoking prevalence and education level among U.S. adults. Using the 2024 Behavioral Risk Factor Surveillance System (BRFSS) dataset, we simulate plausible relationships between these two variables under varying biological and sampling conditions. The ultimate goal is to assess how statistical test performance — particularly power — varies across different effect sizes, noise levels, and sample sizes.

# Objective and Approach

## Specific Objective

**Research Question:** Is smoking prevalence associated with education level among adults in the United States? Specifically, does the probability of being a current smoker decrease as educational attainment increases?

## Data Description

```{r data-load, results='hide'}
# Load data
brfss <- read_xpt("C:/Users/HUAWEI/Desktop/MPH 2026 spring/6270/cp03/LLCP2024XPT/LLCP2024.XPT")

# Select and filter relevant variables
brfss_clean <- brfss %>%
  dplyr::select(`_RFSMOK3`, EDUCA) %>%
  dplyr::filter(
    !is.na(`_RFSMOK3`),
    !is.na(EDUCA),
    EDUCA %in% 1:6,
    `_RFSMOK3` %in% c(1, 2)
  ) %>%

# Recode variables: smoking status as binary, education as three-level factor
  mutate(
    current_smoker = ifelse(`_RFSMOK3` == 2, 1, 0),
    edu_group = case_when(
      EDUCA %in% 1:3 ~ "Low",
      EDUCA == 4     ~ "Medium",
      EDUCA %in% 5:6 ~ "High"
    ),
    edu_group = factor(edu_group, levels = c("Low", "Medium", "High"))
  )
```

```{r variable-table, echo=FALSE}
variable_info <- data.frame(
  `Variable Name` = c("_RFSMOK3", "EDUCA", "current_smoker", "edu_group"),
  Description = c(
    "Current smoking status (derived)",
    "Highest education level completed",
    "Binary smoker indicator (1=Yes, 0=No)",
    "Grouped education (Low/Medium/High)"
  ),
  `Variable Type` = c("Categorical", "Categorical (ordinal)", "Binary", "Categorical (ordinal)"),
  `R Class` = c("numeric", "numeric", "numeric", "factor"),
  check.names = FALSE
)
kable(variable_info, caption = "Overview of Selected Variables")

```

## Data Visualization

```{r data-viz, fig.height=4, fig.width=6, fig.cap="Smoking prevalence decreases with higher educational attainment."}
smoke_by_edu <- brfss_clean %>%
  group_by(edu_group) %>%
  summarise(
    n = n(),
    smoking_prev = mean(current_smoker, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(smoke_by_edu, aes(x = edu_group, y = smoking_prev, fill = edu_group)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(smoking_prev * 100, 1), "%")),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Low" = "#d73027", "Medium" = "#fc8d59", "High" = "#4575b4")) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 0.35)) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Smoking Prevalence by Education Level",
    x = "Education Level",
    y = "Smoking Prevalence"
  )
```

## Description of Plausible Relationships

The bar chart reveals a clear inverse association between educational attainment and smoking prevalence. Individuals with low education have the highest smoking prevalence (22.7%), followed by those with medium education (16.5%), and those with high education show the lowest prevalence (8.1%). This pattern suggests a monotonically decreasing relationship: as education level increases, the probability of being a current smoker decreases. The relationship can be described as a difference in proportions across ordered groups, where each step up in education is associated with a reduction in smoking probability of approximately 7–8 percentage points.

## Parameters of Interest

```{r param-table, echo=FALSE}
# Estimate real observed probabilities for reference
p_obs <- smoke_by_edu$smoking_prev
names(p_obs) <- as.character(smoke_by_edu$edu_group)

param_table <- data.frame(
  Parameter = c(
    "Baseline smoking probability (High education)",
    "Effect size: Low vs High group difference",
    "Effect size: Medium vs High group difference",
    "Noise (individual-level variability)"
  ),
  Symbol = c("p_high", "delta_low", "delta_med", "epsilon"),
  Description = c(
    paste0("Probability of smoking in high education group (observed ≈ ", round(p_obs["High"], 3), ")"),
    "Absolute difference in smoking probability: p_low - p_high",
    "Absolute difference in smoking probability: p_med - p_high",
    "Standard deviation of added Gaussian noise before Bernoulli draw"
  )
)
kable(param_table, caption = "Parameters of Interest for Simulation")
```

# Simulation

## Simulation Basis

The simulation generates a dataset with three education groups (Low, Medium, High), where group membership is assigned randomly based on the observed proportions in the BRFSS data. Smoking status is then drawn as a Bernoulli outcome, where the probability of smoking depends on the group and the specified effect size. Gaussian noise is added to the latent probability before thresholding to introduce individual-level variability.

## Simulation Function

```{r sim-function}
# Observed proportions of each education group from real data
edu_props <- brfss_clean %>%
  count(edu_group) %>%
  mutate(prop = n / sum(n)) %>%
  arrange(edu_group)

simulate_smoking <- function(n,
                              effect_size,   # delta_low = absolute diff low vs high
                              noise,         # SD of Gaussian noise added to latent prob
                              p_high = 0.10, # baseline prob for High education group
                              edu_proportions = c(0.15, 0.30, 0.55)) {
  # Assign education groups
  edu_group <- sample(
    c("Low", "Medium", "High"),
    size = n,
    replace = TRUE,
    prob = edu_proportions
  )

  # Define group-level smoking probabilities
  p_low  <- p_high + effect_size
  p_med  <- p_high + effect_size / 2

  base_prob <- ifelse(edu_group == "Low", p_low,
               ifelse(edu_group == "Medium", p_med, p_high))

  # Add noise on latent scale and clamp to [0,1]
  latent_prob <- base_prob + rnorm(n, mean = 0, sd = noise)
  latent_prob <- pmax(0, pmin(1, latent_prob))

  # Draw Bernoulli outcome
  smoker <- rbinom(n, size = 1, prob = latent_prob)

  # Build data frame
  sim_data <- data.frame(
    edu_group = factor(edu_group, levels = c("Low", "Medium", "High")),
    smoker = smoker
  )

  # Compute summary statistics
  summary_stats <- sim_data %>%
    group_by(edu_group) %>%
    summarise(
      n = n(),
      smoking_prev = mean(smoker),
      .groups = "drop"
    )

  return(list(data = sim_data, summary = summary_stats))
}
```

## Simulation Output Example

```{r sim-example}
set.seed(42)
example_run <- simulate_smoking(n = 1000, effect_size = 0.15, noise = 0.05)
kable(example_run$summary,
      caption = "Example Simulation Output (n=1000, effect=0.15, noise=0.05)",
      digits = 3)
```

## Simulation Automation

```{r sim-automation}
set.seed(123)

# Parameter grid
effect_sizes <- seq(0.02, 0.20, length.out = 10)   # 10 effect sizes
sample_sizes  <- c(100, 250, 500, 1000, 2500,
                   5000, 10000, 50000, 100000, 200000)  # 10 sample sizes (incl. ~BRFSS scale)
noise_levels  <- c(0.02, 0.08, 0.15)               # 3 noise levels: low, medium, high

# Store all results
sim_results <- list()
idx <- 1

for (es in effect_sizes) {
  for (n in sample_sizes) {
    for (noise in noise_levels) {
      result <- simulate_smoking(n = n, effect_size = es, noise = noise)
      sim_results[[idx]] <- list(
        effect_size = es,
        sample_size = n,
        noise       = noise,
        data        = result$data,
        summary     = result$summary
      )
      idx <- idx + 1
    }
  }
}

cat("Total simulation runs:", length(sim_results), "\n")
```

# Visualization

```{r sim-viz, fig.width=10, fig.height=7, fig.cap="Simulated smoking prevalence difference (Low vs High education) across effect sizes and sample sizes, faceted by noise level."}

# Extract summary results into a flat data frame
summary_df <- lapply(sim_results, function(x) {
  prev_low  <- x$summary$smoking_prev[x$summary$edu_group == "Low"]
  prev_high <- x$summary$smoking_prev[x$summary$edu_group == "High"]
  if (length(prev_low) == 0 || length(prev_high) == 0) return(NULL)
  data.frame(
    effect_size    = x$effect_size,
    sample_size    = x$sample_size,
    noise          = x$noise,
    observed_diff  = prev_low - prev_high
  )
}) %>%
  bind_rows()

summary_df$noise_label <- factor(
  paste0("Noise = ", summary_df$noise),
  levels = paste0("Noise = ", sort(unique(summary_df$noise)))
)

ggplot(summary_df,
       aes(x = factor(round(effect_size, 3)),
           y = log10(sample_size),
           fill = observed_diff)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(
    low = "white", mid = "#fee090", high = "#d73027",
    midpoint = 0.1,
    name = "Observed\nPrev. Diff\n(Low - High)"
  ) +
  facet_wrap(~ noise_label, ncol = 3) +
  scale_y_continuous(
    breaks = log10(c(100, 1000, 10000, 100000)),
    labels = c("100", "1,000", "10,000", "100,000")
  ) +
  theme_minimal(base_size = 11) +
  labs(
    title    = "Simulated Smoking Prevalence Difference (Low vs High Education)",
    subtitle = "Across effect sizes, sample sizes, and noise levels",
    x        = "True Effect Size (Absolute Difference in Probability)",
    y        = "Sample Size (log scale)"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text  = element_text(face = "bold")
  )
```

# Interpretation

The simulations highlight three main findings. First, sample size is the most important factor: at n = 100, estimates are highly unstable and unreliable, while at n = 10,000 and above, the observed difference consistently reflects the true effect. Second, larger effect sizes are easier to detect across all conditions, as shown by the consistent left-to-right darkening in the heatmap. Third, noise level has relatively little impact on the results under this setup. Overall, these simulations confirm that the large sample size of the BRFSS dataset provides sufficient power to detect even small differences in smoking prevalence across education groups.

# AI Use Disclosure Statement

This document was prepared with assistance from Claude. Claude was used to generate the initial R Markdown structure, develop the simulation function, and produce visualization code. All outputs were reviewed and interpreted by the author.
